Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique

### 2017-10-30 inst/rmd/ihmt_presentation_brew_2017.Rmd
minor changes

---
title: "Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique"
subtitle: 'IHMT, October, 2017, Lisboa, Portugal'
fig_height: 5
fig_width: 7
bibliography: bibliography.bib
output:
  beamer_presentation:
    includes:
      in_header: preamble_beamer.sty
    theme: "Berlin"
    colortheme: "seagull"
    fonttheme: "structuresmallcapsserif"
---

@@

# Create model data for all obs (not just propensity score)
model_data <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs) %>%
  # left_join(matched) %>%
  mutate(days_since = break_days_since(days_since)) %>%
  filter(!is.na(days_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick)) #%>%
  # # remove the period 30 days after spraying
  # filter(days_since >= 30 | days_since <= 0)
# Need to add seasonality
# Need to adjust for houses on and off site

@@

- Malaria is associated with absenteeism in workers [@Nonvignon_2016].
- Malaria has a negative effect on GDP [@Orem_2012] and growth [@McCarthy_2000].
- Malaria control is cost-effective from the societal/public perspective [@Purdy_2013].
- Indoor residual spraying (IRS) is cost-effective [@Howard_2017], [@White_2011].

@@

## Descriptive: sick absenteeism by time from/to intervention

```{r}
ggplot(data = x,
       aes(x = months_since,
           y = sick_absenteeism_rate,
           group = before)) +
  geom_smooth() +
  geom_step(color = 'darkred', alpha = 0.8) +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) +
  labs(x = 'Months relative to IRS',
       y = 'Absenteeism rate',
       title = 'Before/after IRS',
       subtitle = 'Sick absences only') +
  theme_maragra()
```

## Descriptive: absenteeism by time from/to intervention (by time_period)

@@

## Descriptive: sick absenteeism by time from/to intervention (by time_period)

```{r}

ggplot(data = x %>% mutate(time_period = paste0(time_period)),
       aes(x = months_since,
           y = sick_absenteeism_rate,
           group = before)) +
  geom_smooth() +
  geom_step(color = 'darkred', alpha = 0.8) +
  facet_wrap(~time_period, ncol = 1) +
  geom_vline(xintercept = 0, lty = 2, alpha = 0.5) +
  labs(x = 'Months relative to IRS',
       y = 'Absenteeism rate',
       subtitle = 'Sick absences only') +
  theme_maragra()
```

## Descriptive: absenteeism in binned time period

```{r}
x <- model_data %>%
  group_by(days_since) %>%
  summarise(absences = length(which(absent)),
            eligibles = length(absent)) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)

ggplot(data = x,
       aes(x = days_since,
           y = absenteeism_rate)) +
  geom_bar(stat = 'identity',
           alpha = 0.6,
           fill = 'darkblue') +
  theme_maragra() +
  labs(x = 'Days since IRS',
       y = 'Absenteeism rate',
       title = 'Raw absenteeism as a function of days since IRS') +
  geom_label(aes(label = paste0(round(absenteeism_rate, digits = 2), '%')))
```

## Descriptive: absenteeism in binned time periods by season

```{r}
x <- model_data %>%
  group_by(days_since, time_period) %>%
  summarise(absences = length(which(absent)),
            eligibles = length(absent)) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)

ggplot(data = x %>% mutate(time_period = paste0(time_period)),
       aes(x = days_since,
           y = absenteeism_rate)) +
  geom_bar(stat = 'identity',
           alpha = 0.6,
           fill = 'darkblue') +
  theme_maragra() +
  labs(x = 'Days since IRS',
       y = 'Absenteeism rate',
       title = 'Raw absenteeism as a function of days since IRS by time_period') +
  # geom_label(aes(label = paste0(round(absenteeism_rate, digits = 2), '%'))) +
  facet_wrap(~time_period, ncol = 1) +
  theme(axis.text.x = element_text(angle = 90))
```

## Modeling after matching {.allowframebreaks}

@@

@@

@@

Sick absence only:

```{r}
kable(ors_propensity_sick) 
```

@@

Sick absenteeism only:

```{r}
kable(ors_sick) 
```

@@

- 30-50% reduction in absenteeism in the 6 months after IRS, when administered during malaria season.

@@

- Reduction is large and significant.
- Depending on detailed cost data (pending), IRS may be effective even from a purely financial point of view.

>>>>

---
title: "Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique"
subtitle: 'IHMT, October, 2017, Lisboa, Portugal economicsofmalaria.com/ihmt'
fig_height: 5
fig_width: 7
bibliography: bibliography.bib
output:
  beamer_presentation:
    includes:
      in_header: preamble_beamer.sty
    theme: "Berlin"
    colortheme: "seagull"
    fonttheme: "structuresmallcapsserif"
---

@@

# Create model data for all obs (not just propensity score)
model_data <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs) %>%
  # left_join(matched) %>%
  mutate(days_since = break_days_since(days_since)) %>%
  filter(!is.na(days_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick)) %>%
  mutate(month = format(date, '%b'))
  # # remove the period 30 days after spraying
  # filter(days_since >= 30 | days_since <= 0)
# Need to add seasonality
# Need to adjust for houses on and off site

@@

- Malaria is associated with absenteeism in workers [@Nonvignon_2016].
- Malaria has a negative effect on GDP [@Orem_2012] and growth [@McCarthy_2000].
- Malaria control is cost-effective from the societal/public perspective [@Purdy_2013].
- Indoor residual spraying (IRS) is cost-effective [@Howard_2017], [@White_2011] _from a public perspective_.

@@

<!-- ## Descriptive: sick absenteeism by time from/to intervention -->

<!-- ```{r} -->
<!-- ggplot(data = x, -->
<!--        aes(x = months_since, -->
<!--            y = sick_absenteeism_rate, -->
<!--            group = before)) + -->
<!--   geom_smooth() + -->
<!--   geom_step(color = 'darkred', alpha = 0.8) + -->
<!--   geom_vline(xintercept = 0, lty = 2, alpha = 0.5) + -->
<!--   labs(x = 'Months relative to IRS', -->
<!--        y = 'Absenteeism rate', -->
<!--        title = 'Before/after IRS', -->
<!--        subtitle = 'Sick absences only') + -->
<!--   theme_maragra() -->
<!-- ``` -->

## Descriptive: absenteeism by time from/to intervention (by time period)

@@

<!-- ## Descriptive: sick absenteeism by time from/to intervention (by time_period) -->

<!-- ```{r} -->

<!-- ggplot(data = x %>% mutate(time_period = paste0(time_period)), -->
<!--        aes(x = months_since, -->
<!--            y = sick_absenteeism_rate, -->
<!--            group = before)) + -->
<!--   geom_smooth() + -->
<!--   geom_step(color = 'darkred', alpha = 0.8) + -->
<!--   facet_wrap(~time_period, ncol = 1) + -->
<!--   geom_vline(xintercept = 0, lty = 2, alpha = 0.5) + -->
<!--   labs(x = 'Months relative to IRS', -->
<!--        y = 'Absenteeism rate', -->
<!--        subtitle = 'Sick absences only') + -->
<!--   theme_maragra() -->
<!-- ``` -->

<!-- ## Descriptive: absenteeism in binned time period -->

<!-- ```{r} -->
<!-- x <- model_data %>% -->
<!--   group_by(days_since) %>% -->
<!--   summarise(absences = length(which(absent)), -->
<!--             eligibles = length(absent)) %>% -->
<!--   ungroup %>% -->
<!--   mutate(absenteeism_rate = absences / eligibles * 100) -->

<!-- ggplot(data = x, -->
<!--        aes(x = days_since, -->
<!--            y = absenteeism_rate)) + -->
<!--   geom_bar(stat = 'identity', -->
<!--            alpha = 0.6, -->
<!--            fill = 'darkblue') + -->
<!--   theme_maragra() + -->
<!--   labs(x = 'Days since IRS', -->
<!--        y = 'Absenteeism rate', -->
<!--        title = 'Raw absenteeism as a function of days since IRS') + -->
<!--   geom_label(aes(label = paste0(round(absenteeism_rate, digits = 2), '%'))) -->
<!-- ``` -->

<!-- ## Descriptive: absenteeism in binned time periods by season -->

<!-- ```{r} -->
<!-- x <- model_data %>% -->
<!--   group_by(days_since, time_period) %>% -->
<!--   summarise(absences = length(which(absent)), -->
<!--             eligibles = length(absent)) %>% -->
<!--   ungroup %>% -->
<!--   mutate(absenteeism_rate = absences / eligibles * 100) -->

<!-- ggplot(data = x %>% mutate(time_period = paste0(time_period)), -->
<!--        aes(x = days_since, -->
<!--            y = absenteeism_rate)) + -->
<!--   geom_bar(stat = 'identity', -->
<!--            alpha = 0.6, -->
<!--            fill = 'darkblue') + -->
<!--   theme_maragra() + -->
<!--   labs(x = 'Days since IRS', -->
<!--        y = 'Absenteeism rate', -->
<!--        title = 'Raw absenteeism as a function of days since IRS by time_period') + -->
<!--   # geom_label(aes(label = paste0(round(absenteeism_rate, digits = 2), '%'))) + -->
<!--   facet_wrap(~time_period, ncol = 1) + -->
<!--   theme(axis.text.x = element_text(angle = 90)) -->
<!-- ``` -->

## Modeling after matching I

@@

## Modeling after matching II

@@

## Modeling after matching III

@@

<!-- \vfill -->
<!-- \newpage -->

<!-- Sick absence only: -->

<!-- \begin{tiny} -->

<!-- ```{r} -->
<!-- kable(ors_propensity_sick)  -->
<!-- ``` -->

<!-- \end{tiny} -->

@@

<!-- \vfill -->
<!-- \newpage -->

<!-- Sick absenteeism only: -->

<!-- \begin{tiny} -->

<!-- ```{r} -->
<!-- kable(ors_sick)  -->
<!-- ``` -->

<!-- \end{tiny} -->

@@

- 30-50% reduction in absenteeism in the 6 months after IRS, when administered during malaria season. 
- Next steps are incorporation of (a) productivity data (via cane cut), (b) better clinical data (via local health facilities), and (c) full cost data.  
- Will also be comparing with a sugarcane facility in a zone where an elimination campaign is taking place.

@@

- Reduction is large and significant.
- Depending on detailed cost data (pending), IRS may be effective even from a purely financial point of view (ie, beyong just "corporate social responsibility").
